Modeling $q_{95}$ Windows for the Suppression of Edge Localized Modes by Resonant Magnetic Perturbations in the DIII-D Tokamak

### 2020-09-19 ExperimentalData/DIIID/145380/Paper/paper1.tex
Revised version of Paper1

\title {Modelling $q_{95}$ Windows for the Suppression of Edge Localized Modes by Resonant Magnetic Perturbations in the DIII-D Tokamak}

@@

\begin{abstract}
A toroidal asymptotic matching model of the response of a tokamak plasma to a static resonant magnetic perturbation (RMP) is used to simulate the $n=3$ RMP-induced 
edge-localized-mode (ELM)-suppression
windows in $q_{95}$ that are evident when the plasma current is slowly ramped in DIII-D discharge \#145380. All quantities employed in the simulation are
derived from experimental measurements, apart from the neutral particle data. 
Three cases are considered.
In the first case, the natural frequencies of tearing modes resonant in the plasma are 
determined by the ion flows at the corresponding rational surfaces, which is the prediction of nonlinear theory.
In the second case, the natural frequencies are 
determined by the guiding-center flows at the rational surfaces.
 In the third case, the natural frequencies   are 
determined by the electron flows at the  rational surfaces, which is the prediction of linear theory.
The second case gives  the best agreement between the simulations and the  experimental observations.   The third case gives by far the worst agreement between the simulations 
and the  observations. The first case only leads to partial agreement between  the  simulations and the  observations. However, this lack of complete agreement
may be a consequence of using an inaccurate assumption for the neutral particle distribution in the pedestal.  
\end{abstract}

@@

The application of a static RMP, resonant in the pedestal region (i.e., the region of strong pressure and current density gradients characteristic of the edge region of an H-mode 
tokamak discharge), to an H-mode tokamak discharge is observed to give rise to  two distinct phenomena.\cite{schmitz, lanctot,paz1,d158115,paz} The first of these  is the 
so-called ``density pump-out'', which  is characterized by a reduction in the electron number density
in the pedestal region that varies smoothly with the amplitude of the applied RMP,  is (usually) accompanied by a similar, but significantly smaller, reduction
in the electron and ion temperatures, but  is not associated with ELM suppression. The second phenomenon is  ``ELM suppression'' itself, which 
 occurs when the amplitude of the applied RMP exceeds a certain threshold value. 
ELM suppression  is only observed to take place when $q_{95}$ (i.e., the safety-factor on the magnetic flux-surface that encloses 95\% of the poloidal flux enclosed by
the last closed flux-surface) takes values that lie in certain narrow windows. \cite{paz1,d158115}

@@

Recently, a toroidal generalization of the 
cylindrical asymptotic matching model presented
in Ref.~\onlinecite{rf1} was formulated and
used to model  RMP-induced ELM-suppression experiments performed on the DIII-D tokamak,\cite{rf2}
leading to similar conclusions to the aforementioned TM1
studies. The primary aim of this paper is to employ this
new model to try to account for the  $q_{95}$
ELM-suppression windows  that are apparent when the edge safety-factor is slowly ramped in
a particular DIII-D discharge (\#145380) in which an $n=3$
RMP is used to control ELMs.\cite{d3d,d3d2}

It is well-known that magnetic reconnection driven when a  (stable) tearing mode interacts with a static RMP that is resonant at a particular magnetic flux-surface in a tokamak plasma is
facilitated when the associated ``natural frequency'' is relatively small.\cite{rfa,rfb}  The natural frequency  of a  (stable) tearing mode is the
helical phase velocity that the mode would possess were it naturally unstable (in the absence of the RMP).  Driven magnetic reconnection leads to the
formation of a locked magnetic island chain at the resonant surface in question. 
Hence, the prevailing hypothesis is that a $q_{95}$ window for
RMP-induced ELM suppression occurs when $q_{95}$ is such that the natural frequency of a tearing mode resonant  at 
the top of the pedestal is close to zero.\cite{d3d} 
However, there is currently some uncertainty in the fusion community regarding the form of the natural frequency.

According to nonlinear tearing mode theory, a tearing mode is essentially convected by
the local ion fluid at the rational surface (see the discussion in Sect.~\ref{natural}).\cite{nl1,nl2,nl3}
In fact, if the response of an H-mode tokamak plasma to an applied RMP
is governed by nonlinear physics then we would expect the
natural frequency to take the form\,\cite{rf2}
\begin{equation}\label{e2}
\varpi_{\perp\,i} = -n\left(\omega_E +\left[1-L_{00}^{\,ii}+L_{01}^{\,ii}\left(\frac{\eta_i}{1+\eta_i}\right)\right]
\omega_{\ast\,i}
-\left[L_{00}^{\,iI}- L_{01}^{\,iI}\left(\frac{\eta_I}{1+\eta_I}\right)\right]\omega_{\ast\,I}\right),
\end{equation}
where
\begin{align}
\omega_E({\mit\Psi}_p) &= -\frac{d{\mit\Phi}}{d{\mit\Psi}_p},\\[0.5ex]
\omega_{\ast\,a}({\mit\Psi}_p) &=-\frac{T_a}{Z_a\,e}\,\frac{d\ln p_a}{d{\mit\Psi}_p},\\[0.5ex]
\eta_a({\mit\Psi}_p)&=\frac{d\ln T_a}{d\ln n_a},
\end{align}
for $a=i, I$. 
Here,  $e$ is the magnitude of the electron charge, ${\mit\Psi}_p$  the equilibrium poloidal magnetic
flux (divided by $2\pi$), ${\mit\Phi}({\mit\Psi}_p)$ the equilibrium scalar
electric potential, and $n$ the toroidal mode number of the RMP.  Moreover, $Z_i$, $n_i({\mit\Psi}_p)$, $T_i({\mit\Psi}_p)$, and $p_i({\mit\Psi}_p)=n_i\,T_i$   are the charge number, 
equilibrium number density, 
equilibrium temperature, and equilibrium pressure
of the majority (thermal) ions, respectively,
whereas  $Z_I$, $n_I$, $T_I$, $p_I = n_I\,T_I$
are the corresponding quantities for the impurity ions.  
Furthermore, $L_{00}^{ii}({\mit\Psi}_p)$, $L_{01}^{ii}({\mit\Psi}_p)$,  $L_{00}^{II}({\mit\Psi}_p)$, and $L_{01}^{II}({\mit\Psi}_p)$ are neoclassical parameters that are defined
in Sect.~B of Ref.~\onlinecite{rf2}. Note that these parameters are
affected by charge exchange  with neutrals. 
The right-hand side of Eq.~(\ref{e2}) is evaluated
at the ``rational''  (i.e., resonant) magnetic flux-surface at which the
safety-factor
\begin{equation}
q({\mit\Psi}_p)= \frac{d{\mit\Psi}_t}{d{\mit\Psi}_p},
\end{equation}
takes the rational value $m/n$, where $m$ is a positive integer. Here, $m$ and $n$ are the numbers of poloidal and
toroidal periods, respectively, of the helical magnetic island chain driven at the rational surface. Moreover, ${\mit\Psi}_t({\mit\Psi}_p)$ is the equilibrium toroidal magnetic flux (divided by $2\pi$).

A second possibility is that a tearing mode is convected by the local guiding-center fluid at the rational surface  (again, see the discussion in Sect.~\ref{natural}),\cite{heyn,paz1}
in which case we would expect the natural frequency to take the form 
\begin{equation}\label{e3}
\varpi_{\perp\, EB} =-n\,\omega_E.
\end{equation}
As before, the right-hand side of Eq.~(\ref{e3}) is evaluated
at the rational magnetic flux-surface.

Finally, according to
linear tearing mode theory, a tearing mode is essentially convected by
the local electron fluid at the resonant surface (again, see the discussion in Sect.~\ref{natural}).\cite{hender,cole} Hence, if the response of an H-mode tokamak plasma to an applied RMP
is governed by linear physics then we would expect the
natural frequency to take the form\,\cite{lin1,lin2,lin3}
\begin{equation}
\varpi_{\perp\,e} = - n\,(\omega_E+\omega_{\ast\,e}),\label{e1}
\end{equation}
where
\begin{align}
\omega_{\ast\,e}({\mit\Psi}_p) &=\frac{T_e}{e}\,\frac{d\ln p_e}{d{\mit\Psi}_p}.
\end{align}
 Here, $p_e({\mit\Psi}_p)$ is the equilibrium
electron pressure, and $T_e({\mit\Psi}_p)$ the equilibrium
electron temperature.
Finally,
as before, the right-hand side of Eq.~(\ref{e1}) is evaluated
at the rational magnetic flux-surface.

The secondary aim of this paper is to determine which of the three aforementioned choices for the natural frequency can best account for the   $q_{95}$ 
ELM-suppression windows  observed in DIII-D discharge \#145380.

\section{Description of Theoretical Model}
The theoretical model of the response of a tokamak plasma to an externally applied RMP that is used in this
paper is described in detail in Ref.~\onlinecite{rf2}. The model employs a standard asymptotic matching approach.\cite{fkr,am1,am2,am3}
According to this approach, the response of the plasma to the applied RMP is governed by a combination of flux-freezing and
perturbed force balance (this combination is often referred to as ``marginally-stable ideal-MHD", which is a misnomer because MHD {\em per se}\/ plays no
role) everywhere in the plasma apart from a number of relatively narrow (in the radial direction) regions in which the applied
perturbation resonates with the equilibrium magnetic field. Magnetic reconnection can take place within the resonant regions to
produce relatively thin magnetic islands. Within the resonant regions, the plasma response is governed by nonlinear,
as opposed to linear, two-fluid resistive-MHD. This is the case because the widths of the magnetic island chains
driven at the resonant surfaces exceed the linear layer widths (which invalidates linear theory).\cite{rf1}
Thus, when employing the asymptotic matching approach, the equations of flux-freezing and perturbed force balance
are solved in the so-called ``outer region'' that comprises most of the plasma (and the surrounding vacuum), the equations of 
nonlinear two-fluid resistive-MHD are solved in the various resonant layers that constitute the so-called ``inner region'', and the two
sets of solutions are then asymptotically matched to one another.

A toroidal tokamak equilibrium exhibits two distinct types of response to an applied RMP.\cite{paz1,kink0,kink1}
The first of these is known as the ``tearing response''---this is a non-ideal-MHD response that is associated with the
formation of current sheets and magnetic island chains at various resonant surfaces within the plasma. The second
response type is known as the ``kink response''---this is an edge-localized ideal-MHD response that is associated
with coupling to a stable non-resonant kink mode. For the case of the tearing response, our model employs an
approximation in which the plasma response is assumed to be vacuum-like between the various resonant surfaces. On the other hand,
the kink response of the plasma is calculated exactly using the {\tt GPEC} code.\cite{ipec,gpec} (See Sect.~A.8 of Ref.~\onlinecite{rf2}.)

Our model has been implemented in the newly-developed {\tt EPEC} (Extended Perturbed Equilibrium Code) code. The name
of this code reflects the fact that the nonlinear evolution of tearing modes in a
tokamak plasma has far more in common with the $1\tfrac{1}{2}$-D
evolution of the global plasma equilibrium than it does with
conventional linear tearing mode physics. (See the discussion in Sect.~\ref{general}.) In particular, the Alfv\'{e}n time is
an irrelevant timescale in nonlinear tearing mode theory, and
is also  very much shorter than the timescales on which
physical quantities of interest actually evolve. (Note that all timescales are normalized to the Alfv\'{e}n time
in Ref.~\onlinecite{rf2}. However, this is just a matter of convention. With the benefit of hindsight, it would
have been better to normalize the timescales with respect to a diamagnetic timescale, in which case the Alfv\'{e}n time would have completely dropped
out of the final system of equations. See Sect.~E in Ref.~\onlinecite{rf2}.)

\section{EPEC Modeling of DIII-D Discharge \#145380}
\subsection{Plasma Equilibrium}
DIII-D discharge \#145380 is an ITER-Similar-Shape (ISS), ELMing, H-mode discharge, with a toroidal magnetic field $B_T=-1.9\,{\rm T}$, in which the
plasma current, $I_p$, is slowly ramped over a 2 second interval in order to scan the magnetic safety-factor.\cite{d3d,d3d2} The majority ions are Deuterium, whereas the
minority ions are Carbon~VI.
Figure~\ref{fig1} gives an overview of DIII-D discharge \#145380. 
A static RMP is applied to the plasma by running steady $n=3$ currents through the I-coil system.\cite{icoil}
Three windows of ELM suppression or mitigation are evident: the first (in which ELMs are mitigated, but not
entirely suppressed) extends from $t=2840$--$2980$ ms; the
second (in which ELM are entirely suppressed) extends from $t=3320$--$3560$ ms; and the third (in which ELMs are
entirely suppressed) extends from $t=3880$--$4200$ ms.

Figure~\ref{fig2} shows the measured plasma equilibrium at the start of the current ramp ($t=2500$ ms). Likewise,
Fig.~\ref{fig3} shows the measured plasma profiles at the start of the current ramp. Here, ${\mit\Psi}_N$ is a
normalized equilibrium poloidal magnetic flux defined such that the magnetic axis corresponds to ${\mit\Psi}_N=0$, 
and the last closed magnetic flux-surface (LCFS) to ${\mit\Psi}_N=1$. The perpendicular
diffusivity data comes from {\tt TRANSP} modelling.\cite{transp}

Measured plasma equilibrium and
plasma profiles are reconstructed every $50$ ms. The equilibrium and profiles  at a given point in time are interpolated from
the reconstructed equilibria and profiles. In fact, all quantities that depend on the plasma equilibrium and profiles are recalculated every 1 ms.  Data values from the
{\tt GPEC} code are calculated every 100 ms. As before, the {\tt GPEC} data values at a given point in time are interpolated from the
calculated {\tt GPEC} data values, and any quantity that depends on these values is recalculated every 1 ms. There is one major exception to
the aforementioned scheme. The theoretical model used in this paper actually requires plasma equilibria and profiles that are unaffected
by the applied RMP. However, there is clear evidence that the equilibria and the profiles (in particular) are significantly modified
within the three ELM-suppression/mitigation windows (presumably because wide magnetic island chains are driven in the pedestal). [See Fig.~1(c).]
Hence, we do not use the equilibrium, profile, or {\tt GPEC}, data within the three ELM-suppression/mitigation windows. Instead, we
interpolate across these windows using data taken on either side of the windows.

Figure~\ref{fig4} shows the safety-factor in DIII-D discharge \#145380 as a function of time. It can be seen
that $q_{95}\equiv q({\mit\Psi}_N=0.95)$ varies from about 4.1 at the start of the current ramp to about 3.3
at the end. However, the safety-factor at the magnetic axis, $q_0 \equiv q({\mit\Psi}_N=0.0)$, remains approximately
constant at about 1.0. For the sake of practicality,  the {\tt GPEC} and {\tt EPEC} calculations described in this paper ignore $n=3$ rational
surfaces that lie very close to the LCFS: in other words,  surfaces that lie in the region $0.997 < {\mit\Psi}_N\leq 1.000$. Let $\overline{q_{95}}\equiv 5.2060-0.4457\,t$ represent the
least-squares linear fit to the $q_{95}$ versus time data shown in Fig.~\ref{fig4}. 
It follows that the 
first, second, and third ELM-suppression/mitigation windows extend over the $\overline{q_{95}}$ ranges $3.88$--$3.94$, $3.62$--$3.73$,
and $3.33$--$3.48$, respectively.

Note that all quantities used in our theoretical model are derived from experimental measurements taken in DIII-D discharge \#145380, with the exception of the neutral particle data. 
Unfortunately, no neutral particle data is available for DIII-D discharge \#145380.
Hence, we employ neutral particle data derived from a much earlier discharge (see Sect.~III.D of Ref.~\onlinecite{rf2}).\cite{neutral}
In future, data from the new {\tt LLAMA} (Lyman Alpha Measurement Apparatus) system will be routinely available in DIII-D discharges.\cite{llama}
Combining this data  with modeling from a neutral particle simulation code such as, for example, the  SOLPS-ITER code\,\cite{solps} ought to generate sufficient information to fully determine the
neutral particle data required by the {\tt EPEC} code.

@@

\subsection{Case 2: $\varpi=\varpi_{\perp\,EB}$}
The second case that we shall consider is that in which the natural frequency is determined by the local guiding-center fluid; 
in other words, the natural frequency is specified by Eq.~(\ref{e3}).

>>>>

\title {Modeling $q_{95}$ Windows for the Suppression of Edge Localized Modes by Resonant Magnetic Perturbations in the DIII-D Tokamak}

@@

\begin{abstract}
A toroidal asymptotic matching model of the response of a tokamak plasma to a static resonant magnetic perturbation (RMP) is used to simulate the $n=3$ RMP-induced 
edge-localized-mode (ELM)-suppression
windows in $q_{95}$ that are evident when the plasma current is slowly ramped in DIII-D discharge \#145380. All quantities employed in the simulation are
derived from experimental measurements, apart from the neutral particle data. 
Three cases are considered.
In the first case, the natural frequencies of tearing modes resonant in the plasma are 
determined by the ion flows at the corresponding resonant surfaces, which is the prediction of nonlinear tearing mode theory.
In the second case, the natural frequencies are 
determined by the local ${\bf E}\times {\bf B}$ velocities at the resonant surfaces.
 In the third case, the natural frequencies   are 
determined by the electron flows at the  resonant surfaces, which is the prediction of linear tearing mode theory.
The second case gives  the best agreement between the simulations and the  experimental observations.    The first and third cases only leads to partial agreement between  the  simulations and the  observations. In the first case, the lack of complete agreement
may be a consequence of using an inaccurate assumption for the neutral particle distribution in the pedestal.  In the third case, the lack of complete agreement is probably
due to the fact that the response of a tokamak plasma to an RMP is not accurately described by linear tearing mode theory.
\end{abstract}

@@

The application of a static RMP, resonant in the pedestal region (i.e., the region of strong pressure and current density gradients characteristic of the edge region of an H-mode 
tokamak discharge), to an H-mode tokamak discharge is observed to give rise to  two distinct phenomena.\cite{schmitz, lanctot,paz1,d158115,lyons,paz} The first of these  is the 
so-called ``density pump-out'', which  is characterized by a reduction in the electron number density
in the pedestal region that varies smoothly with the amplitude of the applied RMP,  is (usually) accompanied by a similar, but significantly smaller, reduction
in the electron and ion temperatures, but  is not associated with ELM suppression. The second phenomenon is  ``ELM suppression'' itself, which 
 occurs when the amplitude of the applied RMP exceeds a certain threshold value. 
ELM suppression  is only observed to take place when $q_{95}$ (i.e., the safety-factor on the magnetic flux-surface that encloses 95\% of the poloidal flux enclosed by
the last closed flux-surface) takes values that lie in certain narrow windows. \cite{paz1,d158115}

@@

Recently, a toroidal generalization of the 
cylindrical asymptotic matching model presented
in Ref.~\onlinecite{rf1} was formulated and
used to model  RMP-induced ELM-suppression experiments performed on the DIII-D tokamak,\cite{rf2}
leading to similar conclusions to the aforementioned TM1
studies. The first aim of this paper is to employ this
new model to try to account for the  $q_{95}$
ELM-suppression/mitigation windows  that are apparent when the edge safety-factor is slowly ramped in
a particular DIII-D discharge (\#145380) in which an $n=3$
RMP is used to control ELMs.\cite{d3d,d3d2} The second aim is to determine  which choice for the so-called ``natural frequency''  of tearing modes can best account for the   $q_{95}$ 
ELM-suppression/mitigation windows  observed in DIII-D discharge \#145380.

\section{Natural Frequency}
\subsection{Introduction}
It is well-known that magnetic reconnection driven when a  (stable) tearing mode interacts with a static RMP that is resonant at a particular magnetic flux-surface in a tokamak plasma is
facilitated when the associated natural frequency, in the absence of the RMP,  is relatively small.\cite{rfa,rfb}  The natural frequency  of a  (stable) tearing mode is the
helical phase velocity that the mode would possess were it naturally unstable.  (It should be noted that the maximum practical currents
that can be driven in edge-resonant RMP coils severely limit the ability of such coils to modify the rotation
frequencies of tearing modes resonant in the pedestals of present-day tokamaks.  For example, in the DIII-D tokamak, as will become clear later on in the paper,
the RMP coils are only capable of arresting the rotation of edge-resonant tearing modes when the magnitude of the helical
phase velocities of such modes fall below about 5 krad/s. A typical phase velocity is 100 krad/s.) Driven magnetic reconnection leads to the
formation of a locked magnetic island chain at the resonant surface in question. 
Hence, the prevailing hypothesis is that a $q_{95}$ window for
RMP-induced ELM suppression occurs when $q_{95}$ is such that the natural frequency of a tearing mode resonant  at 
the top of the pedestal is close to zero.\cite{d3d} 
However, there is currently some debate in the fusion community regarding the appropriate choice for the natural frequency.

 The natural frequency of a tearing mode 
is determined by the equilibrium plasma flow at the
resonant surface. Now, a magnetic island chain is a helical
pattern in the magnetic field  generated by a helical current perturbation that is localized in the
vicinity of the resonant surface.
Given that plasma current is predominately carried by the
electrons, it is natural to suppose that a magnetic
island chain (as well as the tearing mode perturbation
away from the resonant surface)  is convected by the
electron fluid in the immediate vicinity of the resonant
surface. This is indeed the case in the linear regime.\cite{ara} Of course, as a consequence of diamagnetic flows, if the island chain is
convected by the electron fluid at the resonant surface then it propagates with respect to the local ion fluid. However, this is not a
problem because a linear layer is sufficiently thin that the magnetic field can diffuse through the plasma very rapidly, which implies that
the ion fluid is not tied to the magnetic structure of the island chain.

The situation is very different in the nonlinear regime. 
The region inside the magnetic separatrix of a nonlinear magnetic island chain is governed by a combination of flux-freezing and perturbed force balance.
This implies that both the electron and the ion fluids are trapped inside the separatrix, and are, therefore,  forced to co-rotate with the island chain. There
is no such constraint outside the separatrix, so the electron and ion fluids rotate at different speeds in this region,
as a consequence of diamagnetism.  It follows that one or other of the electron and the ion fluid rotation profiles must exhibit a strong gradient across
the separatrix. The island propagation velocity is determined by which of the two fluids is most resistant to the formation of such a gradient. 
Of course, it is the ion fluid which is more resistant because of its much greater perpendicular viscosity,\cite{nl1,wat} as well as its much larger 
neoclassical stress tensor.\cite{nl2} Hence, a nonlinear magnetic island chain is convected by the ion fluid in the vicinity of the resonant 
surface, because this choice of propagation speed minimizes the ion fluid velocity gradient across the separatrix.

Finally, we could imagine that if the width of an island chain is neither much less than the linear layer width (which is the strict
criterion for the validity of linear theory) nor much greater than the linear layer width (which is the strict criterion for
the validity of nonlinear theory) then the chain lies in some sort of intermediate regime in which it is
convected at the local ${\bf E}\times {\bf B}$ velocity (this prediction is generally intermediate between the predictions of linear and nonlinear theory).

It is clear, from the preceding discussion,  that there are three main choices for the natural frequency of a given tearing mode in a tokamak plasma.

\subsection{Case 1}
As we have already mentioned, according to nonlinear tearing mode theory, a tearing mode is essentially convected by
the local ion fluid at the resonant surface.\cite{nl1,nl2,nl3}
In fact, if the response of an H-mode tokamak plasma to an applied RMP
is governed by nonlinear physics then we would expect the
natural frequency to take the form\,\cite{rf2}
\begin{equation}\label{e2}
\varpi_{\perp\,i} = -n\left(\omega_E +\left[1-L_{00}^{\,ii}+L_{01}^{\,ii}\left(\frac{\eta_i}{1+\eta_i}\right)\right]
\omega_{\ast\,i}
-\left[L_{00}^{\,iI}- L_{01}^{\,iI}\left(\frac{\eta_I}{1+\eta_I}\right)\right]\omega_{\ast\,I}\right),
\end{equation}
where
\begin{align}
\omega_E({\mit\Psi}_p) &= -\frac{d{\mit\Phi}}{d{\mit\Psi}_p},\\[0.5ex]
\omega_{\ast\,a}({\mit\Psi}_p) &=-\frac{T_a}{Z_a\,e}\,\frac{d\ln p_a}{d{\mit\Psi}_p},\\[0.5ex]
\eta_a({\mit\Psi}_p)&=\frac{d\ln T_a}{d\ln n_a},
\end{align}
for $a=i, I$. 
Here,  $e$ is the magnitude of the electron charge, ${\mit\Psi}_p$  the equilibrium poloidal magnetic
flux (divided by $2\pi$), ${\mit\Phi}({\mit\Psi}_p)$ the equilibrium scalar
electric potential, and $n$ the toroidal mode number of the RMP.  Moreover, $Z_i$, $n_i({\mit\Psi}_p)$, $T_i({\mit\Psi}_p)$, and $p_i({\mit\Psi}_p)=n_i\,T_i$   are the charge number, 
equilibrium number density, 
equilibrium temperature, and equilibrium pressure
of the majority (thermal) ions, respectively,
whereas  $Z_I$, $n_I$, $T_I$, $p_I = n_I\,T_I$
are the corresponding quantities for the impurity ions.  
Furthermore, $L_{00}^{ii}({\mit\Psi}_p)$, $L_{01}^{ii}({\mit\Psi}_p)$,  $L_{00}^{II}({\mit\Psi}_p)$, and $L_{01}^{II}({\mit\Psi}_p)$ are neoclassical parameters that are defined
in Sect.~B of Ref.~\onlinecite{rf2}. Note that these parameters are
affected by charge exchange  with neutrals. 
The right-hand side of Eq.~(\ref{e2}) is evaluated
at the ``rational''  (i.e., resonant) magnetic flux-surface at which the
safety-factor, 
\begin{equation}
q({\mit\Psi}_p)= \frac{d{\mit\Psi}_t}{d{\mit\Psi}_p},
\end{equation}
takes the rational value $m/n$, where $m$ is a positive integer. Here, $m$ and $n$ are the numbers of poloidal and
toroidal periods, respectively, of the helical magnetic island chain driven at the rational surface. Moreover, ${\mit\Psi}_t({\mit\Psi}_p)$ is the equilibrium toroidal (poloidal) magnetic flux (divided by $2\pi$).

\subsection{Case 2}
A second possibility is that a tearing mode is convected at the local ${\bf E}\times{\bf B}$ velocity at the resonant surface,\cite{heyn,lyons,paz1}
in which case we would expect the natural frequency to take the form 
\begin{equation}\label{e3}
\varpi_{\perp\, EB} =-n\,\omega_E.
\end{equation}
As before, the right-hand side of Eq.~(\ref{e3}) is evaluated
at the rational magnetic flux-surface.

\subsection{Case 3}
Finally, according to
linear tearing mode theory, a tearing mode is essentially convected by
the local electron fluid at the resonant surface.\cite{hender,cole} Hence, if the response of an H-mode tokamak plasma to an applied RMP
is governed by linear physics then we would expect the
natural frequency to take the form\,\cite{lin1,lin2,lin3}
\begin{equation}
\varpi_{\perp\,e} = - n\,(\omega_E+\omega_{\ast\,e}),\label{e1}
\end{equation}
where
\begin{align}
\omega_{\ast\,e}({\mit\Psi}_p) &=\frac{T_e}{e}\,\frac{d\ln p_e}{d{\mit\Psi}_p}.
\end{align}
 Here, $p_e({\mit\Psi}_p)$ is the equilibrium
electron pressure, and $T_e({\mit\Psi}_p)$ the equilibrium
electron temperature.
Finally,
as before, the right-hand side of Eq.~(\ref{e1}) is evaluated
at the rational magnetic flux-surface.

\section{EPEC Modeling of DIII-D Discharge \#145380}
\subsection{EPEC Code}
The theoretical model of the response of a tokamak plasma to an externally applied RMP that is used in this
paper is described in detail in Ref.~\onlinecite{rf2}. The model employs a standard asymptotic matching approach.\cite{fkr,coppi,ruth,ara,pletz,am1,pletz1,tokuda,brennan,fura,am2,am3}
According to this approach, the response of the plasma to the applied RMP is governed by a combination of flux-freezing and
perturbed force balance (this combination is often referred to as ``marginally-stable ideal-MHD") everywhere in the plasma apart from a number of relatively narrow (in the radial direction) regions in which the applied
perturbation resonates with the equilibrium magnetic field. Magnetic reconnection can take place within the resonant regions to
produce relatively thin magnetic islands. Within the resonant regions, the plasma response is governed by nonlinear,
as opposed to linear, two-fluid resistive-MHD. This is the case because the widths of the magnetic island chains
driven at the resonant surfaces (in DIII-D discharge 158115,\cite{rf1} which is similar to DIII-D discharge 145380) exceed the linear layer widths (which invalidates linear theory).
(This state of affairs can be expected to persist in future tokamaks, such as ITER, because such tokamaks will operate with much higher values of the
Lundquist number than present-day tokamaks, leading to a reduction in the linear layer widths.)
Thus, when employing the asymptotic matching approach, the equations of flux-freezing and perturbed force balance
are solved in the so-called ``outer region'' that comprises most of the plasma (and the surrounding vacuum), the equations of 
nonlinear two-fluid resistive-MHD are solved in the various resonant layers that constitute the so-called ``inner region'', and the two
sets of solutions are then asymptotically matched to one another.

A toroidal tokamak equilibrium exhibits two distinct types of response to an applied RMP.\cite{paz,kink0,kink1}
The first of these is known as the ``tearing response''---this is a non-ideal-MHD response that is associated with the
formation of current sheets and magnetic island chains at various resonant surfaces within the plasma. The second
response type is known as the ``kink response''---this is an edge-localized ideal-MHD response that is associated
with coupling to a stable non-resonant kink mode. For the case of the tearing response, our model employs an
approximation in which the plasma response is assumed to be vacuum-like between the various resonant surfaces. On the other hand,
the kink response of the plasma is calculated exactly using the {\tt GPEC} code.\cite{ipec,gpec} (See Sect.~A.8 of Ref.~\onlinecite{rf2}.)

Our model has been implemented in the newly-developed {\tt EPEC} (Extended Perturbed Equilibrium Code) code. The name
of this code reflects the fact that the nonlinear evolution of tearing modes in a
tokamak plasma has far more in common with the $1\tfrac{1}{2}$-D
evolution of the global plasma equilibrium than it does with
conventional linear tearing mode physics. In particular, the Alfv\'{e}n time is
an irrelevant  timescale in nonlinear tearing mode theory, and
is also  very much shorter than the timescales on which
physical quantities of interest actually evolve. (Note that all timescales are normalized to the Alfv\'{e}n time
in Ref.~\onlinecite{rf2}. However, this is just a matter of convention. With the benefit of hindsight, it would
have been better to normalize the timescales with respect to a diamagnetic timescale, in which case the Alfv\'{e}n time would have completely dropped
out of the final system of equations. See Sect.~E in Ref.~\onlinecite{rf2}.)

\subsection{Plasma Equilibrium}
DIII-D discharge \#145380 is an ITER-Similar-Shape (ISS), ELMing, H-mode discharge, with a toroidal magnetic field $B_T=-1.9\,{\rm T}$, in which the
plasma current, $I_p$, is slowly ramped over a 2 second interval in order to scan the magnetic safety-factor.\cite{d3d,d3d2} The majority ions are Deuterium, whereas the
minority ions are Carbon~VI.
Figure~\ref{fig1} gives an overview of DIII-D discharge \#145380. 
A static RMP is applied to the plasma by running steady $n=3$ currents through the I-coil system.\cite{icoil}
Three windows of ELM suppression or mitigation are evident: the first (in which ELMs are mitigated, but not
entirely suppressed) extends from $t=2840$--$2980$ ms; the
second (in which ELM are entirely suppressed) extends from $t=3320$--$3560$ ms; and the third (in which ELMs are
entirely suppressed) extends from $t=3880$--$4200$ ms.

Figure~\ref{fig2} shows the measured plasma equilibrium at the start of the current ramp ($t=2500$ ms). Likewise,
Fig.~\ref{fig3} shows the measured plasma profiles at the start of the current ramp. Here, ${\mit\Psi}_N$ is a
normalized equilibrium poloidal magnetic flux defined such that the magnetic axis corresponds to ${\mit\Psi}_N=0$, 
and the last closed magnetic flux-surface (LCFS) to ${\mit\Psi}_N=1$. The perpendicular
diffusivity data comes from {\tt TRANSP} modeling.\cite{transp}

Measured plasma equilibrium and
plasma profiles are reconstructed every $50$ ms. The equilibrium and profiles  at a given point in time are interpolated from
the reconstructed equilibria and profiles. In fact, all quantities that depend on the plasma equilibrium and profiles are recalculated every 1 ms.  Data values from the
{\tt GPEC} code are calculated every 100 ms. As before, the {\tt GPEC} data values at a given point in time are interpolated from the
calculated {\tt GPEC} data values, and any quantity that depends on these values is recalculated every 1 ms. There is one major exception to
the aforementioned scheme. The theoretical model implemented in the {\tt EPEC} code actually requires  profiles that are unaffected
by the applied RMP (to be more exact, it requires sufficient information to reconstruct  natural frequencies that are unmodified by the electromagnetic
torques that develop in the plasma in response to the RMP). However, there is clear evidence that the  profiles  are significantly modified
within the three ELM-suppression/mitigation windows (presumably because wide magnetic island chains are driven in the pedestal). [See Fig.~1(c).]
Hence, we do not use the  profile (or {\tt GPEC}) data within the three ELM-suppression/mitigation windows. Instead, we
interpolate across these windows using data taken on either side of the windows. The ultimate justification for this approach is the fact that, as can be seen from Figs.~\ref{fig5}, \ref{fig8}, and
\ref{fig11}, {\tt EPEC} simulations performed in the absence of the RMP (i.e., in the absence of RMP-generated electromagnetic torques) give very similar natural frequencies to simulations performed in the presence of the RMP, except
in time intervals in which one or more tearing modes lock to the RMP (of course, we are identifying such intervals with the ELM suppression/mitigation windows).

Figure~\ref{fig4} shows the safety-factor in DIII-D discharge \#145380 as a function of time. It can be seen
that $q_{95}\equiv q({\mit\Psi}_N=0.95)$ varies from about 4.1 at the start of the current ramp to about 3.3
at the end. However, the safety-factor at the magnetic axis, $q_0 \equiv q({\mit\Psi}_N=0.0)$, remains approximately
constant at about 1.0. For the sake of practicality,  the {\tt GPEC} and {\tt EPEC} calculations described in this paper ignore $n=3$ rational
surfaces that lie very close to the LCFS; in other words,  surfaces that lie in the region $0.997 < {\mit\Psi}_N\leq 1.000$. (It should be
noted that {\tt EPEC}, like {\tt GPEC}, employs a straight field-line flux coordinate system, and is effectively
spectral in the poloidal and toroidal directions.)  $\overline{q_{95}}\equiv 5.2060-0.4457\,t$ is the
least-squares linear fit to the $q_{95}$ versus time data shown in Fig.~\ref{fig4}. 
It follows that the 
first, second, and third ELM-suppression/mitigation windows extend over the $\overline{q_{95}}$ ranges $3.88$--$3.94$, $3.62$--$3.73$,
and $3.33$--$3.48$, respectively.

Note that all quantities used in our theoretical model are derived from experimental measurements taken in DIII-D discharge \#145380, with the exception of the neutral particle data. 
Unfortunately, no neutral particle data is available for DIII-D discharge \#145380.
Hence, we employ neutral particle data derived from a much earlier discharge (see Sect.~III.D of Ref.~\onlinecite{rf2}).\cite{neutral}
In future, data from the new {\tt LLAMA} (Lyman Alpha Measurement Apparatus) system will be routinely available in DIII-D discharges.\cite{llama}
Combining this data  with modeling from a neutral particle simulation code, such as, for example, the  SOLPS-ITER code,\cite{solps} ought to generate sufficient information to fully determine the
neutral particle data required by the {\tt EPEC} code.

@@

\subsection{Case 2: $\varpi=\varpi_{\perp\,EB}$}
The second case that we shall consider is that in which the natural frequency is determined by the local ${\bf E}\times {\bf B}$ velocity; 
in other words, the natural frequency is specified by Eq.~(\ref{e3}).
